---
title: "LongDat tutorial (longdat_cont)"
subtitle: "Longitudinal analysis (time as continuous variable) with longdat_cont()"
output: 
  rmarkdown::html_vignette:
    toc: true 
vignette: >
  %\VignetteIndexEntry{LongDat_tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
This is an example of running longdat_cont. Note that the time variable here should be continuous If time variable is discrete, please apply longdat_disc instead. 

## Input format

The input format should be the same as the example data set "Fasting_cont.txt".
```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  rmarkdown.html_vignette.check_title = FALSE
)
```

```{r message=FALSE}
# Load the package
library(LongDat)
library(tidyverse)
library(kableExtra)
```

```{r tidy=TRUE, tidy.opts=list(width.cutoff=80)}
# Get the path of example dataset "Fasting_cont.txt" so that we can run LongDat on it 
system.file("Fasting_cont.txt", package = "LongDat")

# If the command above doesn't work, you can also try the one below. It will show you the path to your R libraries on your computer.
.libPaths()
# Copy the path from .libPaths() and add "/LongDat/Fasting_cont.txt" to the end. This is the path to the example dataset "Fasting_cont.txt".

# In addition the methods above, if you have the source code of LongDat package, you can find "Fasting_cont.txt" in the folder "inst/" as wel.
```
First, let's have a look at Fasting_cont.txt. This is a mock longitudinal data set with 2 time points (0 and 7 days). Here we want to see if the fasting intervention has significant effect on gut bacterial abundance or not. 
```{r message=FALSE, tidy=TRUE, tidy.opts=list(width.cutoff=80)}
# Read in Fasting_cont.txt
library(kableExtra)
fasting <- read.table("/Library/Frameworks/R.framework/Versions/4.0/Resources/library/LongDat/Fasting_cont.txt", header = T, row.names = NULL, sep = "\t")
fasting %>%
  kbl() %>%
  kable_paper(bootstrap_options = "responsive", font_size = 12) %>% 
  scroll_box(width = "700px", height = "200px")
```
As you can see, the "Individual" is at the first column, and the dependent variable, which are bacterial abundances in this case, are at the end of the table. Any column apart from individual, test_var (e.g. day) and dependent variables will be taken as potential confounders. For example, here the potential confounders are sex, age, drug A and drug B. For the details of the requirements, please read the help page of this function. **Please avoid using characters that don't belong to [ASCII printable characters](https://www.w3schools.com/charsets/ref_html_ascii.asp) for the column names in the input data.**

## Run longdat_cont()

Below is an example of running longdat_cont(). The input is the example data Fasting_cont.txt, and the data_type is "count" since the dependent variable (bacterial abundance in this case) is count data. The "test_var" is the independent variable you're testing, and here we're testing the "Day". The variable_col is 7 because the dependent variables start at column 7. And the fac_var states the columns that aren't numerical. Lastly, the output_tag is the name tag you'd like to add to the output. For the details of the arguments, please read the help page of this function. The run below takes less than a minute to complete.

```{r message=FALSE, tidy=TRUE, tidy.opts=list(width.cutoff=80)}
# Run longdat_cont on example data set
# When data_type = count, set seed for reproducible randomized control test  
set.seed(100)
example <-  longdat_cont(input = "/Library/Frameworks/R.framework/Versions/4.0/Resources/library/LongDat/Fasting_cont.txt", data_type = "count", test_var = "Day", variable_col = 7, fac_var = c(1, 3), verbose = TRUE)
```
If you have completed running the function successfully, you'll see the message "Finished successfully!" at the end. The results are stored in list format.

## Results
### Result table

Now, let's have a look at the result table. 
```{r message=FALSE, tidy=TRUE, tidy.opts=list(width.cutoff=80)}
#The first dataframe in the list is the result table
result_table <- example[[1]]
result_table %>%
  kableExtra::kbl() %>%
  kableExtra::kable_paper(bootstrap_options = "responsive", font_size = 12, position = "center") %>% 
  kableExtra::scroll_box(width = "700px")
```
The first and second column show the prevalence and mean abundance of each bacterium. According to the "Signal" column, fasting intervention is a significant predictor for BacteriumA as it shows "OK_nc" (which represents "OK and no confounder"), meaning that there is an effect of time and no potential confounder.

The following column "Effect" describes the trend of dependent variables changes along time. Here we can tell that BacteriumA has a decreasing pattern. From the next column "Effect_size", we know that the effect size is -0.78. The important and the most relevant information for users ends here, which are listed from the first column to "Effect_size". 

Then the following columns contain the details of model test p values ("Null_time_model_q"), the post-hoc test p values (Post.hoc_q). For more detailed info of the columns in the result table, please refer to the documentation of this function. The explanation of each type of "Signal" is listed below.

|Signal|Meaning|Explanation|          
|---|--------|-------------|
|NS| Non-significant | There’s no effect of time.|
|OK_nc |OK and no confounder | There’s an effect of time and there’s no potential confounder.|
|OK_d| OK but doubtful |  There’s an effect of time and there’s no potential confounder, however the confidence interval of the test_var estimate in the model test includes zero, and thus it is doubtful of this signal.|
|OK_sd| OK and strictly deconfounded | There are potential confounders, however there’s an effect of time and it is independent of those of confounders. |
|AD| Ambiguously deconfounded | There are potential confounders, and it isn’t possible to conclude whether the effect is resulted from time or confounders.|
|C| Confounded | There’s an effect of time, but it can be reduced to the confounding effects.|

### Confounder table
Next, let's take a look at the confounder table.

```{r}
#The second dataframe in the list is the confounder table
confound_table <- example[[2]]
confound_table %>%
  kbl() %>%
  kable_paper(bootstrap_options = "responsive", font_size = 12, position = "center") %>% 
  scroll_box(width = "700px")
```
The columns of this confounder table should be grouped by every 3 columns. "Confounder1" is the name of the confounder, while "Confounding_type1" is the confounding type of confounder1, and "Effect_size1" is the effect size of the dependent variable values between different levels of confounder 1.  If there are more than 1 confounders, they will be listed along the rows of each dependent variable. Since there is no confounding effect found in this example (according to the result table), the confounder table is a blank.

## Conclusion
From the result above, we confirm that fasting treatment can solely explain the changes in BacteriumA abundance as there is no confounder found. In other words, fasting induces significant change on BacteriumA.

This tutorial ends here! For more detailed descriptions of this function, please look it up with ```?longdat_cont``` in R.

